---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r ETL Data}
library(dplyr)

unzip("activity.zip")
activity <- read.csv("activity.csv")
# Check structure for date format
str(activity)
# Convert date from factor to date format
activity$date <- as.Date(activity$date)

```


## What is mean total number of steps taken per day?
```{r Steps per day}
# Group by day
steps.day <- activity %>% group_by(date) %>% summarise(DailySteps = sum(steps))

hist(steps.day$DailySteps, 
             col = "blue", 
             xlab = "Steps", 
             ylab = "Frequency", 
             main = "Steps per Day")

# Explore data
steps.day.mean <- mean(steps.day$DailySteps, na.rm = TRUE)
steps.day.median <- median(steps.day$DailySteps, na.rm = TRUE)
abline(v=steps.day.mean, col = "red", lwd = 2, lty = 1 )
abline(v=steps.day.median, col = "green", lwd = 2, lty = 2 )

```

*  The mean total number of steps per day is `r format(signif(steps.day.mean,6),big.mark = ",")` steps, indicated as red line on the histogram.

*  The median total number of steps per day is `r format(signif(steps.day.median,6),big.mark = ",")` steps, indicated by the green line on the histogram.

## What is the average daily activity pattern?
```{r Daily Pattern}
steps.interval <- activity %>% group_by(interval) %>% summarise(AvgSteps = mean(steps, na.rm = TRUE))
with(steps.interval, plot(x = interval, y= AvgSteps, type= "l", xlab= "Daily 5 minute intervals", ylab="Steps"))

# 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
Interval.max.steps <- steps.interval$interval[which.max(steps.interval$AvgSteps)]
abline(v = Interval.max.steps,col = "red", lwd = 2, lty = 2 )
```

The most steps was taken during interval `r Interval.max.steps`, as indicated by the red dotted line in the trend.


## Imputing missing values

There are `r sum(is.na(activity))` missing values, which is a small proportion (`r signif(mean(is.na(activity))*100, 2)`%) of the data.

The missing data would be filled with the **median of `r steps.day.median` steps.**

```{r Impute}
# Create a new dataset that is equal to the original dataset but with the missing data filled in.
activity.new <- activity
# combine the datasets by interval and then update the columns
df <- activity.new %>% merge(activity.new, steps.interval, by.x = "interval", by.y = "interval")

```

## Are there differences in activity patterns between weekdays and weekends?
```{r Weekly Pattern}

```

