---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r ETL Data}
library(dplyr)

unzip("activity.zip")
activity <- read.csv("activity.csv")
# Check structure for date format
# str(activity)
# Convert date from factor to date format
activity$date <- as.Date(activity$date)

```


## What is mean total number of steps taken per day?
```{r Steps per day}
# Group by day
steps.day <- activity %>% 
        group_by(date) %>% 
        summarise(DailySteps = sum(steps))

hist(steps.day$DailySteps, 
             col = "blue", 
             xlab = "Steps", 
             ylab = "Frequency", 
             main = "Steps per Day")

# Explore data
steps.day.mean <- mean(steps.day$DailySteps, na.rm = TRUE)
steps.day.median <- median(steps.day$DailySteps, na.rm = TRUE)
abline(v=steps.day.mean, col = "red", lwd = 2, lty = 1 )
abline(v=steps.day.median, col = "green", lwd = 2, lty = 2 )

```

*  The mean total number of steps per day is `r format(signif(steps.day.mean,6),big.mark = ",")` steps, indicated as red line on the histogram.

*  The median total number of steps per day is `r format(signif(steps.day.median,6),big.mark = ",")` steps, indicated by the green line on the histogram.

## What is the average daily activity pattern?
```{r Daily Pattern}
steps.interval <- activity %>% group_by(interval) %>% summarise(AvgSteps = as.integer(ceiling(mean(steps, na.rm = TRUE))))
with(steps.interval, plot(x = interval, y= AvgSteps, type= "l", xlab= "Daily 5 minute intervals", ylab="Steps"))

# 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
Interval.max.steps <- steps.interval$interval[which.max(steps.interval$AvgSteps)]
abline(v = Interval.max.steps,col = "red", lwd = 2, lty = 2 )
```

The most steps was taken during interval `r Interval.max.steps`, as indicated by the red dotted line in the trend.


## Imputing missing values

There are `r sum(is.na(activity))` missing values, which is a small proportion (`r signif(mean(is.na(activity))*100, 2)`%) of the data.

The missing data would be filled with the average value of the 5-minute interval.

```{r Impute}
# Create a new dataset that is equal to the original dataset but with the missing data filled in.
activity.impute <- activity
# Set the avg steps as integer

# combine the datasets by interval and then update the columns
        activity.impute <- merge(x = activity, 
                                 y = steps.interval, 
                                 all.x = TRUE)

        activity.impute$steps[is.na(activity.impute$steps)] <- activity.impute$AvgSteps

# Histogram
        steps.day.impute <- activity.impute %>% 
                group_by(date) %>% 
                summarise(DailySteps = sum(steps))
        
        hist(steps.day.impute$DailySteps, 
                     col = "blue", 
                     xlab = "Steps", 
                     ylab = "Frequency", 
                     main = "Steps per Day (Imputed)")

steps.day.impute.mean <- mean(steps.day.impute$DailySteps, na.rm = TRUE)
steps.day.impute.median <- median(steps.day.impute$DailySteps, na.rm = TRUE)
abline(v=steps.day.impute.mean, col = "red", lwd = 2, lty = 1 )
abline(v=steps.day.impute.median, col = "green", lwd = 2, lty = 2 )

```

After imputing the missing values with the average of each of the 5-minute intervals, the following changes are observed:

*  The mean total number of steps per day reduced from `r format(signif(steps.day.mean,6),big.mark = ",")` steps to `r format(signif(steps.day.impute.mean,6),big.mark = ",")` steps, as indicated by red line on the histogram.

*  The median total number of steps per day reduced from `r format(signif(steps.day.median,6),big.mark = ",")` steps to `r format(signif(steps.day.impute.median,6),big.mark = ",")` steps, indicated by the green line on the histogram.

The overall effect is visible in the lower values of the distribution in the histogram.

## Are there differences in activity patterns between weekdays and weekends?

```{r Weekly Pattern}
# Add column to show type of day week/weekend
table(weekdays(steps.day.impute$date))
Weekend <- c("Saturday", "Sunday")

activity.impute <- mutate(activity.impute, 
                           DayType = as.factor(
                                   ifelse(weekdays(activity.impute$date) %in% Weekend,
                                            "Weekend", "Week")))

steps.interval.impute <- activity.impute %>% 
        group_by(interval, DayType) %>% 
        summarise(AvgSteps = as.integer(ceiling(mean(steps, na.rm = TRUE))))

library(lattice)
xyplot(steps.interval.impute$AvgSteps ~ steps.interval.impute$interval |
        steps.interval.impute$DayType,
        type = c("l", "smooth"),
        xlab = "Interval (5-minutes)", 
        ylab = "Average Steps")
```

Looks like an desk jockey that gets around on weekends.

